---
title: "Step 9 - The multivariate CAR model with common strength of spatial correlation"
description: "We explore extensions of the univariate CAR model described in Hierarchical Modeling and Analysis for Spatial data."
author:
  - name: Matthew Shisler
    affiliation: North Carloina State University - Department of Statistics
    affiliation-url: https://statistics.sciences.ncsu.edu/ 
date: "3/05/2023"
categories: [Bayesian, MCMC, Spatial, CAR] # self-defined categories
draft: false 
format:
  html: 
    code-fold: true
execute: 
  cache: true
  freeze: false
---


```{r}
#| label: load-packages
#| output: false
#| code-summary: "Code: Load the packages"

library(tidyverse)
library(igraph)
library(viridis)
```

## Intro

Here we will extend the univariate CAR model to the multivariate case. 

Consider a spatial domain $\mathcal{D} \in \mathbb{R}^2$ that is partitioned into $n$ areal units. The structure of the spatial domain is captured in the neighborhood matrix. Recall our definition of a neighborhood matrix $\mathbf{W} = \{w_{ij}\}$, where
$$
w_{ij} = 
\begin{cases}
1 \quad \text{if} \quad j \in \mathcal{N}(i),\\
0 \quad \text{otherwise}.
\end{cases}
$$
and by convention $w_{ii}=0$. Further, define $w_{i+} = \sum_{j=1}^pw_{ij}$, i.e. the number of neighbors of location $i$, $|\mathcal{N}(i)|$.

In the univariate case we specified a spatial random effect $\boldsymbol\phi = (\phi_1,\dots,\phi_n)$ meant to characterize spatial dependence. In the multivariate case, say of dimension $p$, we specify a $p \times 1$ spatial random vector $\boldsymbol\phi_i = (\phi_{i1}, \phi_{i2},\dots, \phi_{ip})$ at each location $i$, $i = 1,\dots,n$. Arrange these vectors as rows in a matrix $\boldsymbol\Phi$
$$
\boldsymbol\Phi = 
\begin{pmatrix}
\boldsymbol\phi_{1}\\
\boldsymbol\phi_2\\
\vdots\\
\boldsymbol\phi_n
\end{pmatrix}
=
\begin{pmatrix}
\phi_{11} & \phi_{12} & \dots & \phi_{1p}\\
\phi_{21} & \phi_{22} & \dots & \phi_{2p}\\
\vdots & \vdots & \ddots & \vdots\\
\phi_{n1} & \phi_{n2} & \dots & \phi_{np}\\
\end{pmatrix}.
$$

Define $\boldsymbol\phi' = \text{vec}(\boldsymbol\Phi^T)$, i.e. stacked columns of $\boldsymbol\Phi^T$. Here we will use $\boldsymbol\phi' = \text{vec}(\boldsymbol\Phi^T)$ to characterize spatial dependence.

We concern ourselves with the joint distribution $\boldsymbol\phi'$ by specifying conditional distributions of $\boldsymbol\phi_i$.
Under the Markov Random Field (MRF) assumption, the conditional distributions of $\boldsymbol\phi_i$ can be specified as
$$
p(\boldsymbol\phi_i|\boldsymbol\phi_{j\ne i}, \boldsymbol\Gamma_i) = N\left(\sum_{i \sim j} \mathbf{B}_{ij}\boldsymbol\phi_j, \boldsymbol\Gamma_i\right), \quad i,j = 1,\dots,n.
$$

where $\boldsymbol\Gamma_i$ and $\mathbf{B}_{ij}$ are $p \times p$ matrices. Here we use $i \sim j$ to say that $j$ is in the neighborhood of $i$, i.e. $j \in \mathcal{N}(i)$.

The role of $\boldsymbol\Gamma_i$ and $\mathbf{B}_{ij}$ are analogous to the roles of $\tau_i^2$ and $b_{ij}$, respectively, in the univariate CAR models. The matrix $\boldsymbol\Gamma_i$ is the within-location covariance matrix which describes the dependence of the variables in the vector $\boldsymbol\phi_i$. The matrix $\mathbf{B}_{ij}$ is a matrix that allows us to weight observations from locations in the neighborhood of location $i$. A convenient special case is to set $\mathbf{B}_{ij} = b_{ij}\mathbf{I}_p$ where $b_{ij} = w_{ij}/w_{i+}$, i.e. the neighboring observations are equally weighted.

Specifying the conditional distributions in this way implies the unique joint distribution, via Brook's Lemma, to be
$$
p(\boldsymbol\phi' \;| \;\{\boldsymbol\Gamma_i\}) \propto \exp\left\{-\frac{1}{2}\boldsymbol\phi^T\boldsymbol\Gamma^{-1}(\mathbf{I}_{np} - \tilde{\mathbf{B}})\phi\right\}
$$
where $\boldsymbol\Gamma$ is block-diagonal with block $\boldsymbol\Gamma_i$, and $\tilde{\mathbf{B}}$ is $np \times np$ with the $(i,j)$-th block $\mathbf{B}_{ij}$.

Of course, the conditions for a proper multivariate normal distribution are symmetry and positive definiteness of $\boldsymbol\Gamma^{-1}(\mathbf{I}_{np} - \tilde{\mathbf{B}})$. 

Setting $\mathbf{B}_{ij} = b_{ij}\mathbf{I}_p$ where $b_{ij} = w_{ij}/w_{i+}$ leads to the symmetry condition $b_{ij}\boldsymbol\Gamma_j = b_{ji}\boldsymbol\Gamma_i$. 

Further, a common simplifying assumption would be to consider equal within-location covariance. That is, set $\boldsymbol\Gamma_i = w_{i+}^{-1}\boldsymbol\Lambda$ where $\boldsymbol\Lambda$ is the $p \times p$ within-location covariance matrix for $\boldsymbol\phi_i$ common across spatial locations $i=1,\dots,n$.

Under these assumptions we can write $\boldsymbol\Gamma = (\mathbf{D}^{-1}\otimes\boldsymbol\Lambda)$ where $\mathbf{D}$ is an $n \times n$ diagonal matrix with $\mathbf{D}_{ii} = w_{i+}$. And also $\tilde{\mathbf{B}} = \mathbf{B}\otimes\mathbf{I}_p$ where $\mathbf{B}$ is a $n \times n$ matrix and elements of $\mathbf{B}$ are $\mathbf{B} = \{b_{ij}\} = \{w_{ij}/w_{i+}\}$, not to be confused with the block matrix specification $\mathbf{B}_{ij}$ given previously. 

With this in mind, we can write the precision of the above joint distribution in an alternate form,

\begin{align*}

\Gamma^{-1}(\mathbf{I}_{np} - \tilde{\mathbf{B}}) &= (\mathbf{D}^{-1} \otimes \boldsymbol\Lambda)^{-1}(\mathbf{I}_{np} - \mathbf{B}\otimes \mathbf{I}_p)\\
&= (\mathbf{D} \otimes \boldsymbol\Lambda^{-1})(\mathbf{I}_{np} - \mathbf{B}\otimes \mathbf{I}_p) & \text{Kronecker Prod - Inverse}\\
&= (\mathbf{D} \otimes \boldsymbol\Lambda^{-1}) - (\mathbf{D} \otimes \boldsymbol\Lambda^{-1})( \mathbf{B}\otimes \mathbf{I}_p) & \text{(Matrix Mult - distributive wrt matrix addition)}\\
&= (\mathbf{D} \otimes \boldsymbol\Lambda^{-1}) - (\mathbf{D}\mathbf{B}) \otimes (\boldsymbol\Lambda^{-1}\mathbf{I}_p) & \text{(Kronecker Prod - mixed-product property)}\\
&= (\mathbf{D} - \mathbf{D}\mathbf{B})\otimes\boldsymbol\Lambda^{-1} & \text{(Kronecker Prod - distributive wrt matrix addition)}\\
&= (\mathbf{D} - \mathbf{W}) \otimes \boldsymbol\Lambda^{-1}
\end{align*}

where the last step follows from the fact that $\mathbf{B} = \mathbf{D}^{-1}\mathbf{W}$. 

Now there is no need to construct a large block-diagonal $\boldsymbol\Gamma$ matrix. Also, the neighborhood matrix $\mathbf{W}$ and $\mathbf{D}$, the diagonal matrix of its row sums, are fixed and known quantities easily computed outside of any MCMC sampling loop. The joint distribution is then

$$
p(\boldsymbol\phi' \;| \boldsymbol\Lambda) \propto \exp\left\{-\frac{1}{2}\boldsymbol\phi^T\left((\mathbf{D} - \mathbf{W})\otimes \boldsymbol\Lambda^{-1}\right)\phi\right\}
$$
or 
$$
\boldsymbol\phi'|\boldsymbol\Lambda \sim N\left(\boldsymbol 0, (\mathbf{D} - \mathbf{W})^{-1} \otimes \boldsymbol\Lambda\right).
$$

However, we again encounter the issue that $(\mathbf{D} - \mathbf{W}) \otimes \boldsymbol\Lambda^{-1}$ is singular, since $(\mathbf{D} - \mathbf{W})$ is singular. This motivates the introduction of a scalar "spatial strength" parameter, $\rho$, also analogous to the univariate case. There are in fact more general conditions for recovering positive definiteness, but these are not explored here. More details can be found in. . .

Introducing $\rho$ results in a proper joint distribution with form

$$
p(\boldsymbol\phi' \;| \boldsymbol\Lambda) \propto \exp\left\{-\frac{1}{2}\boldsymbol\phi^T\left((\mathbf{D} - \rho\mathbf{W})\otimes \boldsymbol\Lambda^{-1}\right)\boldsymbol\phi\right\}
$$
provided that $|\rho|<1$.

Let's simulate a simple scenario. Let the spatial domain be partitioned into a $40 \times 40$ regular lattice with $n = 1600$ spatial locations. The regular lattice partition will induce a neighborhood matrix $\mathbf{W}$ and diagonal matrix $\mathbf{D}$ for the row sums of $\mathbf{W}$. 

Let $p = 2$ so that we have a $p \times p$ vector $\boldsymbol\phi_i$ at each location $i = 1,\dots,n$. Next, specify the common within-location covariance matrix 
$$
\boldsymbol\Lambda =
\begin{pmatrix}
1 & 0.9\\
0.9 & 1 
\end{pmatrix}
$$
This way $\text{Var}(\phi_{i1}) = 1$, $\text{Var}(\phi_{i2}) = 1$, and $\text{Corr}(\phi_{i1}, \phi_{i2}) = 0.9$. Lastly, set $\rho = 0.99$ to induce strong spatial correlation. Here is a naive way to sample from the joint distribution for $\boldsymbol\phi'$. It is naive because it does not take advantage of the sparsity of the precision matrix. We leave computational advantages to a later investigation.

```{r}
set.seed(831)

# specify spatial domain
n <- 40^2
spat_domain <- expand.grid(x = 1:sqrt(n), y = 1:sqrt(n))
spat_domain$label <- 1:n

spat_domain_g <- make_lattice(c(sqrt(n),sqrt(n)), mutual = TRUE)
W <- as_adjacency_matrix(spat_domain_g, sparse=0)
D <- diag(rowSums(W))

# specify parameters
p <- 2
Lambda <- matrix(c(1, 0.9,
                   0.9, 1), byrow = T, ncol = p)
rho <- 0.99

# construct precision matrix for join dist of phi
inv_Sigma <- kronecker((D-rho*W), solve(Lambda))

# sample from joint dist of phi (technique borrowed for precision matrices)
spat_phi <- backsolve(chol(inv_Sigma), rnorm(n*p))

# inverse vectorize phi  
spat_phi <- matrix(spat_phi, byrow = T, ncol = p)

# assign values to spatial data set
spat_domain$phi1 <- spat_phi[,1]
spat_domain$phi2 <- spat_phi[,2]

```

Since we specified this example with both strong spatial and strong within-location correlation, plotting the data for $\phi_{i1}$ and $\phi_{i2}$, we should expect to observe the same spatial correlation patterns for the first and second elements of $\boldsymbol\phi_i$. 

```{r}
#| label: fig-phi1
#| fig-align: "center"
#| fig-cap: "Plot of..."

ggplot(spat_domain) +
  geom_tile(aes(x, y, fill=phi1)) +
  labs(fill = bquote(phi[i1])) +
  scale_y_reverse() +
  scale_fill_gradientn(colors = viridis(10), limits = c(-3.3,3.3)) +
  coord_fixed() + 
  theme_void()

```

```{r}
#| label: fig-phi2
#| fig-align: "center"
#| fig-cap: "Plot of ..."


ggplot(spat_domain) +
  geom_tile(aes(x, y, fill=phi2)) +
  labs(fill = bquote(phi[i2])) +
  scale_y_reverse() +
  scale_fill_gradientn(colors = viridis(10), limits = c(-3.3,3.3)) +
  coord_fixed() + 
  theme_void()
```

Good news! It looks like we were successful in simulating from the MCAR model! For the next post, we will examine the case where $\rho$ is not shared among elements of $\boldsymbol\phi_i$